---
title: "BIEN Data for mapping and biostatistics"
output: html_notebook
---

[BIEN](http://biendata.org)

Today, to build on our previous lesson with BIEN, we will re-visit some of the work we did to be sure proper libraries and some data are loaded. (As you know, no need to re-install.)

```{r}
#install.packages("BIEN")
library("BIEN")
library(ape) #Package for working with phylogenies in R
library(maps) #Useful for making quick maps of occurrences
library(sp) # A package for spatial data
#vignette("BIEN")
```

```{r}
# Download the data for a focal species.
Artocarpus_altilis <- BIEN_occurrence_species(species = "Artocarpus altilis")

# Download the data for California Blackberry (Rubus ursinus).
Rubus_ursinus <- BIEN_occurrence_species(species = "Rubus ursinus")

# Sugarcane (Saccharum spontaneum).
Saccharum_spontaneum <- BIEN_occurrence_species(species = "Saccharum spontaneum")

# California Redwood (Sequoia sempervirens)
Sequoia_sempervirens <- BIEN_occurrence_species(species = "Sequoia sempervirens")
```

It looks like nothing happened, but you actually have created a data frame called Artocarpus_altilis, full of character data and numeric data. Use the commands below. The first allows you to view the so called meta-data (data about the data; how many observations, for example). The second command looks at the first six entries in the data frame. The third looks at the last six entries in the data frame.

```{r}
str(Artocarpus_altilis)
head(Artocarpus_altilis)
tail(Artocarpus_altilis)

# blackberry
str(Rubus_ursinus)
head(Rubus_ursinus)

# sugarcane
str(Saccharum_spontaneum)
head(Saccharum_spontaneum)
```

As you know from looking at the browser, you can look up a rangemap. 
```{r}
breadfruit_rangemap <- BIEN_ranges_load_species("Artocarpus altilis")

# CA blackberry
Rubus_ursinus_rangemap <- BIEN_ranges_load_species("Rubus_ursinus")

# sugarcane
Saccharum_spontaneum_rangemap <- BIEN_ranges_load_species("Saccharum_spontaneum")

# redwood 
Sequoia_sempervirens_rangemap <- BIEN_ranges_load_species("Sequoia_sempervirens")

```

```{r}
# Map the range of breadfruit
map('world', fill=T, col="grey", bg = "light blue", xlim=c(-180,20),ylim=c(-60,80))
plot(breadfruit_rangemap, col="green", add=T)

# CA blackberry
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,20),ylim=c(-60,80))
plot(Rubus_ursinus_rangemap, col="green", add=T)

# redwood 
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,20),ylim=c(-60,80))
plot(Sequoia_sempervirens_rangemap, col="green", add=T)

# sugarcane 
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,20),ylim=c(-60,80))
plot(Saccharum_spontaneum_rangemap, col="green", add=T)
# NOTE: the range map doesnt come up/appears on the map. In my data (listed in "Enviroment" on the right, I see a listing for "Saccharum_spontaneum_rangemap" - so I know it exists (also it runs without error). It just seems that the range map may not exist in the dataset/package or that it wont load - I took about an hour and a half to double check that variable names, quotes, etc. matched up, so I dont think its that. And because the "BIEN_range_get_species" function worked for the first two species, I would think that the error is not in that part of the interface.)
```

```{r}
# Map the range of breadfruit + occurence records
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(breadfruit_rangemap, col="green", add=T)
points(cbind(Artocarpus_altilis$longitude, Artocarpus_altilis$latitude), col="blue", pch=20, cex=1)

# blackberry
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(Rubus_ursinus_rangemap, col="green", add=T)
points(cbind(Rubus_ursinus$longitude, Rubus_ursinus$latitude), col="blue", pch=20, cex=1)

# redwood
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(Sequoia_sempervirens_rangemap, col="green", add=T)
points(cbind(Sequoia_sempervirens$longitude, Sequoia_sempervirens$latitude), col="blue", pch=20, cex=1)

# sugarcane 
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(Saccharum_spontaneum_rangemap, col="green", add=T)
points(cbind(Saccharum_spontaneum$longitude, Saccharum_spontaneum$latitude), col="blue", pch=20, cex=1)
# NOTE: for sugarcane the range map doesnt come up/appears on the map. In my data (listed in "Enviroment" on the right) I see a listing for "Saccharum_spontaneum_rangemap" - so I know it exists/has been created (also it runs without error). It just seems that the range map may not exist in the dataset/package or that it wont load. Since the "BIEN_range_get_species" function worked for the first two species, I would think that the error is not in that part of the interface.
```

By default, RBIEN commands for species occurrence only downloads limited information, but those defaults can be changed to get more expansive data, including:
- species in cultivation are added to those occurring naturally
- occurrence data outside of the Americas
- more taxonomy data (authority, higher levels than genus and species)
- when occurrence happened, country of occurrence, etc.

```{r}
Artocarpus_altilis_full <- BIEN_occurrence_species(species = "Artocarpus altilis",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)

str(Artocarpus_altilis_full)
```

This was a slightly slower process because so much more data were obtained. 

Now let's map again, but on the world map instead of just the Americas. And let's show points from the full query in red, and overlay points from the default query (more limited) in blue. 

```{r}
map('world', fill=T , col="grey", bg="light blue")

#Plot the points from the full query in red
points(cbind(Artocarpus_altilis_full$longitude, Artocarpus_altilis_full$latitude), col="red", pch=20, cex=1)

# Plot the points from the default query in blue
points(cbind(Artocarpus_altilis$longitude, Artocarpus_altilis$latitude), col="blue", pch=20, cex=1) 

# SUGARCANE: SO much more data
Saccharum_spontaneum_full <- BIEN_occurrence_species(species = "Saccharum spontaneum",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)

str(Saccharum_spontaneum_full)

map('world', fill=T , col="grey", bg="light blue")

#Plot the points from the full query in red
points(cbind(Saccharum_spontaneum_full$longitude, Saccharum_spontaneum_full$latitude), col="red", pch=20, cex=1)

# Plot the points from the default query in blue
points(cbind(Saccharum_spontaneum$longitude, Saccharum_spontaneum$latitude), col="blue", pch=20, cex=1) 
```

### Trait Data and linking this to Biostatistics

```{r}
BIEN_trait_list()
```

```{r}
# Create data frame of breadfruit traits
breadfruit_traits <- BIEN_trait_genus(genus = "Artocarpus")
str(breadfruit_traits)
breadfruit_traits

# appple genus traits
apple_traits <- BIEN_trait_genus(genus = "Malus")
str(apple_traits)
apple_traits
```

```{r}
Breadfruit_DBH <- BIEN_trait_traitbygenus(trait = "diameter at breast height (1.3 m)", genus = "Artocarpus")
head(Breadfruit_DBH)
tail(Breadfruit_DBH)
str(Breadfruit_DBH)

# APPLE (diameter at breast height)
apple_DBH <- BIEN_trait_traitbygenus(trait = "diameter at breast height (1.3 m)", genus = "Malus")
head(apple_DBH) # start
tail(apple_DBH) # end
str(apple_DBH)

# APPLE (flower color)
apple_flower_color <- BIEN_trait_traitbygenus(trait = "flower color", genus = "Malus")
head(apple_flower_color) # start
tail(apple_flower_color) # end
str(apple_flower_color)

# APPLE (whole plant height)
apple_WPH <- BIEN_trait_traitbygenus(trait = "whole plant height", genus = "Malus")
head(apple_WPH) # start
tail(apple_WPH) # end
str(apple_WPH)
```

```{r}
# List variable names from Breadfruit_DBH data frame
names(Breadfruit_DBH)

# for apple_DBH
names(apple_DBH)

# for whole plant height of Malus
names(apple_WPH)

```

```{r}
Breadfruit_DBH_small = Breadfruit_DBH[c(1:5)]
Breadfruit_DBH_small
Breadfruit_DBH_small = as.data.frame(Breadfruit_DBH_small)
class(Breadfruit_DBH_small)
head(Breadfruit_DBH_small)
class(Breadfruit_DBH_small$trait_value)
DBH = as.numeric(Breadfruit_DBH_small$trait_value) 
class(DBH)

# APPPLE
apple_DBH_small = apple_DBH[c(1:5)] # only variables 1-5
apple_DBH_small
apple_DBH_small = as.data.frame(apple_DBH_small)
class(apple_DBH_small)
head(apple_DBH_small)
class(apple_DBH_small$trait_value)
aDBH = as.numeric(apple_DBH_small$trait_value)
class(aDBH) # numeric

# APPPLE
apple_WPH_small = apple_WPH[c(1:5)] # only variables 1-5
apple_WPH_small
apple_WPH_small = as.data.frame(apple_WPH_small)
class(apple_WPH_small)
head(apple_WPH_small)
class(apple_WPH_small$trait_value)
aWPH = as.numeric(apple_WPH_small$trait_value)
class(aWPH) # numeric

```

Note that a library is being loaded below, and without it the boxplots will not be generated. This should be a package installed by default when R is installed or updated, but if it's not then you will need to install it.

```{r}
# DBH of BREADFRUIT 
library(lattice)
head(Breadfruit_DBH_small)
boxplot(DBH~scrubbed_species_binomial, data=Breadfruit_DBH_small)

# DBH of Apples
library(lattice)
head(apple_DBH_small)
boxplot(aDBH~scrubbed_species_binomial, data=apple_DBH_small)

# spread of whole plant height of Apples
library(lattice)
head(apple_WPH_small)
boxplot(aWPH~scrubbed_species_binomial, data=apple_WPH_small)


```
  	
```{r}
# whole plant height for BREADFRUIT
Breadfruit_WPH <- BIEN_trait_traitbygenus(trait = "whole plant height", genus = "Artocarpus")
str(Breadfruit_WPH)
Breadfruit_WPH
```

```{r}
names(Breadfruit_WPH)
```

```{r}
Breadfruit_WPH_small = Breadfruit_WPH[c(1:5)]
Breadfruit_WPH_small
Breadfruit_WPH_small = as.data.frame(Breadfruit_WPH_small)
class(Breadfruit_WPH_small)
head(Breadfruit_WPH_small)
class(Breadfruit_WPH_small$trait_value)
WPH = as.numeric(Breadfruit_WPH_small$trait_value)
class(WPH)
```



```{r}
mean(DBH, data=Breadfruit_DBH_small)
mean(WPH, data=Breadfruit_WPH_small)
``` 


Downloading for just one species: 

```{r}
Anitidus_DBH <- BIEN_trait_traitbyspecies(trait = "diameter at breast height (1.3 m)", species = "Artocarpus nitidus")
head(Anitidus_DBH)
tail(Anitidus_DBH)
str(Anitidus_DBH)

# Southern Crabapple (Malus angustifolia) - 169 obs 
Mangustifolia_WPH <- BIEN_trait_traitbyspecies(trait = "whole plant height", species = "Malus angustifolia")
head(Mangustifolia_WPH)
tail(Mangustifolia_WPH)
str(Mangustifolia_WPH)

# Sweet Crabapple (Malus fusca) - 202 obs asiatica
Mcoronaria_WPH <- BIEN_trait_traitbyspecies(trait = "whole plant height", species = "Malus coronaria")
head(Mcoronaria_WPH)
tail(Mcoronaria_WPH)
str(Mcoronaria_WPH)
Mcoronaria_WPH
```


```{r}
Anitidus_DBH_small = Anitidus_DBH[c(1:5)]
Anitidus_DBH_small
Anitidus_DBH_small = as.data.frame(Anitidus_DBH_small)
class(Anitidus_DBH_small)
head(Anitidus_DBH_small)
class(Anitidus_DBH_small$trait_value)
DBHa = as.numeric(Anitidus_DBH_small$trait_value)
class(DBHa)

# angustifolia (for whole plant width) - 
Mangustifolia_WPH_small = Mangustifolia_WPH[c(1:5)]
Mangustifolia_WPH_small
Mangustifolia_WPH_small = as.data.frame(Mangustifolia_WPH_small)
class(Mangustifolia_WPH_small)
head(Mangustifolia_WPH_small)
class(Mangustifolia_WPH_small$trait_value)
WPHangustifolia = as.numeric(Mangustifolia_WPH_small$trait_value)
class(WPHangustifolia)

# coronaria
Mcoronaria_WPH_small = Mcoronaria_WPH[c(1:5)]
Mcoronaria_WPH_small
Mcoronaria_WPH_small = as.data.frame(Mcoronaria_WPH_small)
class(Mcoronaria_WPH_small)
head(Mcoronaria_WPH_small)
class(Mcoronaria_WPH_small$trait_value)
WPHcoronaria = as.numeric(Mcoronaria_WPH_small$trait_value)
class(WPHcoronaria)
```





```{r}
boxplot(DBHa)
hist(DBHa)

```


Another species: 


```{r}
Ahirsutus_DBH <- BIEN_trait_traitbyspecies(trait = "diameter at breast height (1.3 m)", species = "Artocarpus hirsutus")
head(Ahirsutus_DBH)
tail(Ahirsutus_DBH)
str(Ahirsutus_DBH)
```


```{r}
Ahirsutus_DBH_small = Ahirsutus_DBH[c(1:5)]
Ahirsutus_DBH_small
Ahirsutus_DBH_small = as.data.frame(Ahirsutus_DBH_small)
class(Ahirsutus_DBH_small)
head(Ahirsutus_DBH_small)
DBHh = as.numeric(Ahirsutus_DBH_small$trait_value)
class(Ahirsutus_DBH_small$trait_value)
class(DBHh)
```



```{r}
library(dplyr)
Artocarpus2<-rbind(Mcoronaria_DBH_small, Ahirsutus_DBH_small)
head(Artocarpus2)
tail(Artocarpus2)
class(Artocarpus2$trait_value)
DBHboth = as.numeric(Artocarpus2$trait_value)
class(DBHboth)

# putting data togetehr/pooling (for apple)
library(dplyr)
Malus2<-rbind(Mcoronaria_WPH_small, Mangustifolia_WPH_small)
head(Malus2) # shows coronaria @ start
tail(Malus2) # shows agustifolia @ end
class(Malus2$trait_value)
WPHbothMalus = as.numeric(Malus2$trait_value)
class(WPHbothMalus)
```

```{r}
#boxplot(DBHboth)
#hist(DBHboth)

# ASSIGNEMTN: VISUALIZATIONS OF DATA: histograms of each species and combined 
# 2 species from Malus: angustifolia and coronaria
# for whole plant height (WPH)

# angustafolia frequency data (boxplot and histogram)
boxplot(WPHangustifolia)
hist(WPHangustifolia)

# coronaria frequency data (boxplot and histogram)
boxplot(WPHcoronaria)
hist(WPHcoronaria)

# ** x axis to represnt WPH measure

# visual rep of compiled data for 2 species for the DBH
boxplot(WPHbothMalus)
hist(WPHbothMalus)
```

```{r}
ArtocarpusBoth<-cbind(Artocarpus2, DBHboth)
head(ArtocarpusBoth)
tail(ArtocarpusBoth)

# puting data side by side/separate (Malus)
MalusBoth<-cbind(Malus2, WPHbothMalus)
head(MalusBoth)
tail(MalusBoth)
```



```{r}
boxplot(DBHboth~scrubbed_species_binomial, data=ArtocarpusBoth)

# compare WPH data of each species (Malus)
boxplot(WPHbothMalus~scrubbed_species_binomial, data=MalusBoth)

```

```{r}

# plotting plant height and elevation 
plot(x=heights$Midparent, y=heights$Height)

```

Below there is another package being installed into the library and then loaded from the library to be active in the R software. This allows summarizing within species (rather than the whole genus), generating standard functions such as means and standard deviations (sd). The online resource Quick-R can be helpful if you want a bit more support in generating summary statistics, especially within sub-groups. 

https://www.statmethods.net/stats/descriptives.html

```{r}
install.packages("doBy")
library(doBy)
summaryBy(DBH~scrubbed_species_binomial, data=Breadfruit_DBH_small, FUN=c(mean, sd))
summaryBy(WPH~scrubbed_species_binomial, data=Breadfruit_WPH_small, FUN=c(mean, sd))

# mean and sd WPH by species for Manus data
library(doBy)
summaryBy(aWPH~scrubbed_species_binomial, data=apple_WPH_small, FUN=c(mean, sd)) 
```

```{r}
(unique(Breadfruit_DBH_small$scrubbed_species_binomial))
```

```{r}
(unique(Breadfruit_WPH_small$scrubbed_species_binomial))

# species list for manus WPH data
(unique(apple_WPH_small$scrubbed_species_binomial))
```

## Exercises for completing the portfolio 
Do you think the difference in whole plant height is statistically significant between Artocarpus hirsutus and Artocarpus integer?

Can you combine what was learned in the previous lesson "Introduction to getting, testing and plotting data"? 

Can you choose the most appropriate of the following two tests, re-write and run the code, and create the last portion of this R Notebook to include output, and some interpretive comments about statistical support for or against that hypothesis? 

Also try this for a second species pair, which require obtaining some data from BIEN using R BIEN. 

Some possibilities might be: 

cashews are from Anacardium occidentale and perhaps one of the wild cashews (see Wikipedia)

avocados are from Persea americana but there are other relatives in the genus Persea, notably P. coyo

chocolate and one (or a group) of its relatives within Theobroma

choose two species in the genus Agave (used for fiber, tequila; see Wikipedia)

# Population Mean Between Two Independent Samples
# http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-independent-samples

# Population Mean Between Two Matched Samples
# http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-matched-samples

```{r}
# Assigment Code


# MAPS

# blackberry
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(Rubus_ursinus_rangemap, col="green", add=T)
points(cbind(Rubus_ursinus$longitude, Rubus_ursinus$latitude), col="blue", pch=20, cex=1)

# redwood
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(Sequoia_sempervirens_rangemap, col="green", add=T)
points(cbind(Sequoia_sempervirens$longitude, Sequoia_sempervirens$latitude), col="blue", pch=20, cex=1)

# sugarcane 
map('world', fill=T, col="grey", bg="light blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(Saccharum_spontaneum_rangemap, col="green", add=T)
points(cbind(Saccharum_spontaneum$longitude, Saccharum_spontaneum$latitude), col="blue", pch=20, cex=1)
# NOTE: for sugarcane the range map doesnt come up/appears on the map. In my data (listed in "Enviroment" on the right) I see a listing for "Saccharum_spontaneum_rangemap" - so I know it exists/has been created (also it runs without error). It just seems that the range map may not exist in the dataset/package or that it wont load. Since the "BIEN_range_get_species" function worked for the first two species, I would think that the error is not in that part of the interface.

# ----------------
# DATA COMPARISON

# histograms and boxplots of each species and combined 
# 2 species from Malus: angustifolia and coronaria
# for whole plant height (WPH)

# angustafolia frequency data (boxplot and histogram)
boxplot(WPHangustifolia) # same plot that is on the comparison boxplot
hist(WPHangustifolia)

# coronaria frequency data (boxplot and histogram)
boxplot(WPHcoronaria) # same plot that is on the comparison boxplot
hist(WPHcoronaria)

# visual rep of compiled (cumulative) data for 2 species for the DBH
boxplot(WPHbothMalus)
hist(WPHbothMalus)

# ----------------

# box plots for latitude (in 2 selected species)
boxplot(latitude_angustifolia)
boxplot(latitude_coronaria)

# ----------------

# compare WPH data of each species (Malus)
boxplot(WPHbothMalus~scrubbed_species_binomial, data=MalusBoth)

# ----------------

# Scatter plots of latitude vs height in coronaria and angustiolia species
# looks for 1) trend between variables and 2) compariosn of trend between 2 species

# coronaria latitudes vs height
latitude_coronaria = as.numeric(Mcoronaria_WPH$latitude)
WPH_coronaria = as.numeric(Mcoronaria_WPH$trait_value)
# scatter plot (x axis: height // y axis: latitude)
plot(WPH_coronaria, latitude_coronaria)

# plots for latitude and height - coronaria
#boxplot(latitude_coronaria)
#boxplot(WPH_coronaria)

# angustifolia latitudes vs height
latitude_angustifolia = as.numeric(Mangustifolia_WPH$latitude)
WPH_angustifolia = as.numeric(Mangustifolia_WPH$trait_value)
# scatter plot (x axis: height // y axis: latitude)
plot(WPH_angustifolia, latitude_angustifolia)

# ----------------

# mean and sd WPH by species for the genus Manus (comes out as a table)
library(doBy)
summaryBy(aWPH~scrubbed_species_binomial, data=apple_WPH_small, FUN=c(mean, sd)) 



# things that I would like to have been able to do (but didnt have knowledge to):
# - put both scatterplots on one graph
# - better label my graphs 
# - color data points on scatterlots
# - format axis labels 
```



```{r}
# Population Mean Between Two Independent Samples
# http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-independent-samples
ttest_result1 = t.test(x=Breadfruit_WPH_small$scrubbed_species_binomial, y=Breadfruit_WPH_small$trait_value, paired=FALSE, alternative="two.sided")
ttest_result1
```

# This will prepare you for Digital Lab 2

Another feature of BIEN is that it can be limited to specified spatial places. 

One built-in feature is to obtain data for different political jurisdictions such as nations, states or provinces within nations, and counties or parishs with states/provinces.

The line in the chunk of code below can be used to obtain  a comprehensive list of all those political jurisdictions which is useful to figure out how to ask for, say "United States" rather than "USA" and so forth. 

```{r}
BIEN_metadata_list_political_names()
```




```{r}
Cuba<-BIEN_list_country(country = "Cuba")
head(Cuba)
tail(Cuba)
#Panama<-BIEN_list_country(country = "Panama")
#head(Panama)
#tail(Panama)
```


```{r}
Delaware <- BIEN_occurrence_state(country="United States", state="Delaware")
head(Delaware)
tail(Delaware)
```

The rest of the code shows how to get county lists.  

```{r}
#bronx <- BIEN_occurrence_county(country="United States", state="New York", county = "Bronx", native.status=TRUE)
#head(bronx)
#tail(bronx)
OrangeCal_list<-BIEN_list_county(country = "United States", state="California", county="Orange")
head(OrangeCal_list)
tail(OrangeCal_list)

# my 3 counties
MendocinoCal_list<-BIEN_list_county(country = "United States", state="California", county="Mendocino")
head(MendocinoCal_list)
tail(MendocinoCal_list)

KaneUta_list<-BIEN_list_county(country = "United States", state="Utah", county="Kane") # part of which in Zion 
head(KaneUta_list)
tail(KaneUta_list)

NorthSlopeAla_list<-BIEN_list_county(country = "United States", state="Alaska", county="North Slope Borough")
head(NorthSlopeAla_list)
tail(NorthSlopeAla_list)

```


####To complete this portfolio:

Before our next digital lab on Feb 22, choose one of the exercises above and complete it. 

Also use this last coding exercise to obtain species lists for three counties in a state in the USA outside of New York. For example, you could start with the county where you attended high school. Or, you could use a county where you have vacationed in the past, or where you would like to visit. Also download data for two adjacent counties. 
